Debugging by \textit{lastChange}

### 2010-10-22 papers/pldi2011/sigplanconf-template.tex
some improvements to section 3

The story starts when developer examines the program state at a
breakpoint hit and asks for the \textit{lastChange} of a value. The
\textit{lastChange} algorithm is based on program
re-execution. Therefore it is assumed that the execution can be
reproduced automatically (e.g., by a given test case) or with the
developer's assistance. The breakpoint hit becomes the
\textit{reproduction point}. Debugger sets hooks on all instructions
that might be the result of \textit{lastChange} query, re-executes the
program, at every hook hit, \textit{change event}, collects limited
data and once the execution reaches the reproduction point, it
analyzes the collected data and shows the result. The result is a
partial snapshot of the program state at the execution point that last
change occured.

As we shown in the previous section, a \textit{lastChange} query can
be performed on the result of another \textit{lastChange} query. If we
name the reproduction point \textit{R}, we can write the first
\textit{lastChange} in the introductory example in this form:
\textit{lastChange(R, myObject.myProperty)}. It means that this query
is defined at \textit{R}. If we name the result of this query
\textit{A}, we can write the second \textit{lastChange} in this form:
\textit{lastChange(A, myCondition.value)}. In this way, a sequence of
\textit{lastChange} queries with any length can be defined.

\textit{lastChange} can be called on two different types of values: An
object property or a variable value. In JavaScript, every available
variable in a frame comes from one of the scopes in the frame's scope
chain. We explain each case separately.

\subsection{\textit{lastChange} on object propery}
To simplify the algorithm explanation and avoid technical details, we
assume two basic operations and later we explain the details of these
two operations. The first opertion is \texttt{objectId()} which gets a
JavaScript object and returns an integer as its identifier. This
identifier is unique for every object during one execution. The second
operation is \texttt{setPropertyChangeHook()} which gets function as
the callback and a string as the property name like
\texttt{foo}. Whenever property \texttt{foo} in any object changes the
callback will be called. The callback function receives a reference to
the owner of \texttt{foo}.

@@

Once developer asks for the last change of \texttt{bar.foo} at the
reproduction point, debugger calls \texttt{setPropertyChangeHook()}
with \texttt{foo} as the property name and re-executes the
program. Whenever \texttt{foo} changes and the callback function is
called, debugger first calls \texttt{objectId()} on the \texttt{foo}
owner object. Then it stores owner id and stack frame locations. The
reason we keep object id instead of a reference to the object itself
is to let garbage collector to destroy dead objects. Whenever the
execution reaches the reproduction point it looks at the history of
\texttt{foo} changes and finds the last \texttt{foo} change with the
same object id as \texttt{bar} id at the reproduction point. Figure
~\ref{fig:foo-changes1} shows the list of property \texttt{foo} change
events in a hypothetical execution. \texttt{bar} id at the
reproduction point is 1010, so the last change of \texttt{bar.foo} is
the fourth column.

@@

\subsubsection{global and \texttt{with} scopes}
Global scope is the the most outer scope in the scope chain and it is
also refered as global object which is usually \texttt{window}
object. This scope is a regular JavaScript object and therefore every
global variable is a property of global object. Similar to global
scope, \texttt{with} scopes are also regular objects. A \texttt{with}
scope is created by a \texttt{with()} block with an object as the
parameter. Every property of passed object is available inside the
block as a variable. Debugger treats the case where variable's scope
is global or \texttt{with}, like \textit{lastChange} on an object
property.

@@

A variable in these types of scope can only be changed in the function
the variable is defined or the nested fuctions. For example in Figure
~\ref{fig:js-closure}, variable \texttt{var1} in line 11 can only be
changed in lines 10 to 23. Therefore, after locating the function in
which the variable is defined, it is enough to parse the code inside
the function block and set a hook on all lines where the variable is
assigned a new value. % TODO lines or instruction??? 
We also set a hook at the first line of the function which is
corresponding to the line variable is defined. In Figure
~\ref{fig:js-closure}, if the execution is paused at line 17 and the
last change of \texttt{var1} is queried, two hooks on lines 16 and 17
will be set, but if the execution is paused at line 20 , four hooks on
lines 11, 12, 14, 20 will be set. These two cases are different due
the fact that \texttt{var1} in \texttt{firstChild} is a local variable
but in \texttt{secondChild} is a closure variable.

@@

Everytime one of the hooks hits a new change event is inserted to the
list of change events. During execution the same variable might be
changed several times but in different scopes. Figure
~\ref{fig:recursive} demonstrates a case that the same variable
\texttt{x} is changed several times but in different scopes due to the
recursive calls. Without the scope id calling \textit{lastChange(x)}
at point C will return point B while the right answer is point A. To
address this issue, we employ \texttt{scopeId()} operation-explained
in the next section-which returns a unique id for a scopes during on
execution. The scope id is kept at every variable change event. Once
the execution reaches the reproduction point, using variable's scope
id, the last change is recognized.

@@

\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
According to the algorithm explained in the previous section none of
the points B and C can be identified before reaching point
A. Therefore, all \texttt{x} changes for piont B and all \texttt{y}
changes for point C are stored. Once the execution reaches point A,
and the exact \texttt{bar} id reveals, point B can be identified among
the change events in the history of \texttt{x} changes. The problem
arises in recognizing point C. To recognize point C, we need
\texttt{baz} id at point B, but B is a past program state, and
\texttt{baz} id has not been collected at this point.

>>>>

The \textit{lastChange} algorithm is based on program re-execution of
a program halted on a breakpoint. The breakpoint hit becomes the
\textit{reproduction point}. The algorithm starts when developer
examines the program state at a breakpoint hit and asks for the
\textit{lastChange} of a value.  Debugger sets hooks (a callback
function dependent upon the underlying runtime) on all instructions
that might be the result of \textit{lastChange} query. Then the
debugger re-executes the program and every time the hook hits it
checks for a \textit{change event}. In the case of a change, it stores
part of the program state values.  Once the execution reaches the
reproduction point, it analyzes the collected data and shows the
result.  The program state at the execution point for the last change
event is the \textit{lastChange}.

While \textit{lastChange} itself is not a different kind 'breakpoint',
all of the assumptions developers already have for breakpoints hold
for \textit{lastChange}. In particular the re-execution is just the
same operation developers use to debug with breakpoints. Because
\textit{lastChange} dramatically reduces the developers work in
setting and removing breakpoints, the re-execution step will become a
much larege fraction of the debug time. Therefore automatic mechanisms
for re-execution will be corresponding more valuable for debugging
with \textit{lastChange}.

As we described in the preceding section, a \textit{lastChange} query can
be performed on the result of another \textit{lastChange} query. If we
name the reproduction point \textit{R}, we can write the first
\textit{lastChange} in the introductory example in this form:
\textit{lastChange(R, myObject.myProperty)}. It means that this query
is defined at \textit{R}. If we name the result of this query
\textit{A}, we can write the second \textit{lastChange} in this form:
\textit{lastChange(A, myCondition.value)}. In this way, a sequence of
\textit{lastChange} queries with any length can be defined.

\textit{lastChange} can be called on two different types of values: An
object property or a variable value. Three different variations of the
algorithm are required, one for object property and two more depending
upon the scope of a variable.

\subsection{\textit{lastChange} on object propery}
To simplify the algorithm explanation and defer technical details, we
define two basic operations and later we explain the details of these
two operations. The first opertion is \texttt{objectId()}: given a
JavaScript object it returns an integer as its identifier. This
identifier is unique to the object during one execution. The second
operation is \texttt{setPropertyChangeHook()}: given a function and a
string, the function is called whenever a property changes and its
name matches the string For example, if the property might be
\texttt{foo} so changes to \texttt{bar.foo} or \texttt{baz.foo} would
call the function.  The callback function receives a reference to the
owner of \texttt{foo}.

@@

To see how these functions work, suppose the developer asks for the
last change of \texttt{bar.foo} at the reproduction point in a
program. The debugger calls \texttt{setPropertyChangeHook()} with
\texttt{foo} as the property name and re-executes the
program. Whenever \texttt{foo} changes and the callback function is to
be called, debugger first calls \texttt{objectId()} on the
\texttt{foo} owner object. Then it stores this owner id, the stack
frame locations, and other state values in scope at the call point.
Whenever the execution reaches the reproduction point the debugger
looks at the history of \texttt{foo} changes and finds the last
\texttt{foo} change with the same object id as \texttt{bar} id at the
reproduction point. Figure ~\ref{fig:foo-changes1} shows the list of
property \texttt{foo} change events in a hypothetical
execution. \texttt{bar} id at the reproduction point is 1010, so the
last change of \texttt{bar.foo} is the fourth column.  By using an
object id instead of an object reference we allow the garbage
collector to reclaim the space for dead objects just at would in the
absence of the debugger.

@@

\subsubsection{global and \texttt{with} scopes}
Global scope is the the most outer scope in the scope chain and it is
also refered to as the global object (the \texttt{window}
object in Web pages). This scope is a regular JavaScript object and therefore every
global variable is a property of global object. Similarly,
 \texttt{with} scopes are also regular objects. A \texttt{with}
scope is created by a \texttt{with()} block with an object as the
parameter. Every property of this parameter object is available inside the
block as a variable. Debugger treats the case where variable's scope
is global or \texttt{with}, like \textit{lastChange} on an object
property.

@@

Only operations in the scope of the these varibles can change their value; that includes the defining scope and scopes nested inside of them.
For example in Figure
~\ref{fig:js-closure}, variable \texttt{var1} in line 11 can only be
changed in lines 10 to 23. Therefore, after locating the function in
which the variable is defined, it is enough to parse the code inside
the function block and set a hook on all lines where the variable is
assigned a new value. \textit{What if we called this operation setVariableChangeHook() and defered the discussion to 4??}
We also set a hook at the first line of the function which is
corresponding to the line variable is defined. In Figure
~\ref{fig:js-closure}, if the execution is paused at line 17 and the
last change of \texttt{var1} is queried, two hooks on lines 16 and 17
will be set, but if the execution is paused at line 20 , four hooks on
lines 11, 12, 14, 20 will be set. These two cases are different due
the fact that \texttt{var1} in \texttt{firstChild} is a local variable
but in \texttt{secondChild} is a closure variable.

@@

Every time one of the hooks hits a new change event is inserted to the
list of change events. During execution the same variable might be
changed several times but in different scopes. Figure
~\ref{fig:recursive} demonstrates a case that the same variable
\texttt{x} is changed several times but in different scopes due to the
recursive calls. Without the scope id calling \textit{lastChange(x)}
at point C will return point B while the right answer is point A. To
address this issue, we employ \texttt{scopeId()} operation-explained
in the next section-which returns a unique id for a scopes during on
execution. The scope id is kept at every variable change event. Once
the execution reaches the reproduction point, using variable's scope
id, the last change is recognized.

@@

\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
According to the algorithm explained in the previous section none of
the points B and C can be identified before reaching point
A. Therefore, all \texttt{x} changes for point B and all \texttt{y}
changes for point C are stored. Once the execution reaches point A,
and the exact \texttt{bar} id reveals, point B can be identified among
the change events in the history of \texttt{x} changes. The problem
arises in recognizing point C. To recognize point C, we need
\texttt{baz} id at point B, but B is a past program state, and
\texttt{baz} id has not been collected at this point.
